# Names and values {.unnumbered}

```{r setup}
#| include: FALSE

# Set knitr options
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE
)

# Load packages

library(pacman)

p_load(
  lobstr,
  bench,
  here,
  tidyverse
)

# Declare location
here::i_am("quarto/1_chapter2.qmd")

# Set seed
set.seed(6969)

```

# Quiz

## Quiz 1

Given the following data frame, how do I create a new column called "3" that contains the sum of 1 and 2? You may only use `$`, not `[[`. What makes 1, 2, and 3 challenging as variable names?

```{r Quiz 1 Q}
df <- data.frame(runif(3), runif(3))
names(df) <- c(1, 2)
```

### Answer

`1` and `2` and `3` are challenging as variable names since they are non-syntactic names. Though they can still be used (here) via coercing:

```{r Quiz 1 A}
df$`3` <- df$`1` + df$`2`

head(df)

```

## Quiz 2

In the following code, how much memory does y occupy?

```{r Quiz 2 Q}
x <- runif(1e6)
y <- list(x, x, x)

```

### Answer

In this code, y takes slightly more memory than x, but not 3 times more (only 80 bytes). This is because y is a list containing 3 references to the 1 vector underlying x. List structure takes up 80 bytes, the vector takes up 8MB.

```{r Quiz 2 A}
lobstr::obj_size(x)
lobstr::obj_size(y)
lobstr::obj_size(y) - lobstr::obj_size(x)

```

## Quiz 3

On which line does `a` get copied in the following example?

```{r Quiz 3 Q}
a <- c(1, 5, 3, 2)
b <- a
b[[1]] <- 10

```

### Answer

`a` gets copied at line 3 of previous block `b[[1]] <- 10`, since that is when `b` is modified and a new vector is created to contain data that b points to.

# Binding basics

## Ex. 1

Explain the relationship between `a`, `b`, `c`, and `d` in the following code:

### Answers

-   `a` points to a vector
-   `b` points to the same vector as `a`
-   `c` points to the same vector as `a` and `b`
-   `d` points to a different vector

```{r Ex 2.2.1}
a <- 1:10
b <- a
c <- b
d <- 1:10

# Proof
lobstr::obj_addr(a)
lobstr::obj_addr(b)
lobstr::obj_addr(c)
lobstr::obj_addr(d)

```

## Ex. 2

The following code accesses the mean function in multiple ways. Do they all point to the same underlying function object? Verify this with `lobstr::obj_addr()`.

```{r eval = FALSE}
mean
base::mean
get("mean")
evalq(mean)
match.fun("mean")
```

### Answer

All expressions above point to the same function

```{r}
mean %>%
  lobstr::obj_addr()

base::mean %>%
  lobstr::obj_addr()

get("mean") %>%
  lobstr::obj_addr()

evalq(mean) %>%
  lobstr::obj_addr()

match.fun("mean") %>%
  lobstr::obj_addr()

```

## Ex. 3

By default, base R data import functions, like `read.csv()`, will automatically convert non-syntactic names to syntactic ones. Why might this be problematic? What option allows you to suppress this behaviour?

### Answer

`read.csv()` is problematic because they by default creates syntactic names from headers. This changes data without user input.

```{r}
#| eval: false
?read.csv
```

Set `check.names = FALSE` to suppress this behaviour

## Ex. 4

What rules does `make.names()` use to convert non-syntactic names into syntactic ones?

### Answer

```{r}
#| eval: false
?make.names
```

The rules for `make.names()` are:

-   Prepend `X` if first character is invalid
-   Translate all invalid characters to `.`
-   Append `.` if name matches R keyword
-   Deduplicate names

The function itself is written in `C`; I don't know how to understand `C` yet.

```{r}
make.names
```

## Ex. 5

I slightly simplified the rules that govern syntactic names. Why is `.123e1` not a syntactic name? Read `?make.names` for the full details.

### Answer

`.123e1` is not a syntactic name because the remaining text following `.` can be intepreted as a scientific number `123e1 = 1.23`

# Copy-on-modify

## Ex. 1

Why is `tracemem(1:10)` not useful?

### Answer

```{r}
cat(tracemem(1:10), "\n")
cat(tracemem(1:10), "\n")
cat(tracemem(1:10), "\n")
cat(tracemem(1:10), "\n")
```

The purpose of `tracemem()` is to tag a named object and see when its underlying data structure is copied. `1:10` is not a named object, so `tracemem()` wouldn't be useful.

## Ex. 2

Explain why `tracemem()` shows two copies when you run this code. Hint: carefully look at the difference between this code and the code shown earlier in the section.

```{r}
x <- c(1L, 2L, 3L)
tracemem(x)

x[[3]] <- 4
```

### Answer

`tracemem()` shows two copies because `x[[3]]` was replaced with a double. This means the object that x points to was modified (thus copied) twice; first to coerce a integer vector to a double vector, second to replace `x[[3]]` with `4`

```{r}
x <- c(1L, 2L, 3L)
lobstr::tree(x)

x[[3]] <- 4
lobstr::tree(x)
```

## Ex. 3

Sketch out the relationship between the following objects:

```{r}
a <- 1:10
b <- list(a, a)
c <- list(b, a, 1:10)
```

### Answer

Relationship is as follows

-   `a` points to an ALTREP vector representing `1:10`
-   `b` is a list containing 2 references to the ALTREP vector representing `1:10`
-   `c` is a list containing a reference to `b`, a reference to `a`, and a reference to the ALTREP vector representing `1:10`

```{r}
tree(a)

tree(b)

tree(c)
```

## Ex. 4

What happens when you run this code?

```{r}
x <- list(1:10)
x[[2]] <- x
```

### Answer

Originally, `x` is a list with a reference to the vector `1:10`

`x`'s underlying list is then copied-on-modify; then the 2nd slot `x[[2]]` is changed to contain the original list object.

Both list slots point to the same vector `1:10`

```{r}
# First object
x <- list(1:10)
str(x)
tree(x)
ref(x)

# Second object
x[[2]] <- x
str(x)
tree(x)
ref(x)
```

# Object size

## Ex. 1

In the following example, why are `object.size(y)` and `obj_size(y)` so radically different? Consult the documentation of `object.size()`.

```{r}
y <- rep(list(runif(1e4)), 100)

object.size(y)
obj_size(y)

```

### Answer

`object.size(y)` and `obj_size(y)` are different because `object.size(y)` does not take into account the degree of memory sharing in the list.

## Ex. 2

Take the following list. Why is its size somewhat misleading?

```{r}
funs <- list(mean, sd, var)
obj_size(funs)

```

### Answer

The size are misleading because `mean`, `sd`, `var` are base R functions; the list does not contain exact copies, but only references. As in Ex. 1, this would be shown via `object.size()` and `obj_size()`

```{r}
object.size(funs)
obj_size(funs)

```

## Ex. 3

Predict the output of the following code:

```{r echo=TRUE}
a <- runif(1e6)

# This will be 8MB, which is total size of a
obj_size(a)

b <- list(a, a)

# This will be 8MB, which is mostly size of a
obj_size(b)

# This will be 8MB, which is mostly size of a
obj_size(a, b)

b[[1]][[1]] <- 10

# This will be 16MB, which is mostly size of modified b[[1]] and
# unmodified b[[2]] pointing to a
obj_size(b)

# This will be 16MB, which is mostly size of modified b[[1]] and
# unmodified b[[2]] pointing to a
obj_size(a, b)

b[[2]][[1]] <- 10

# This will be 16MB, which is mostly size of modified b[[1]] and
# modified b[[2]]
obj_size(b)

# This will be 24MB, which is mostly size of a, modified b[[1]], and
# modified b[[2]]
obj_size(a, b)

```

# Modify-in-place

## Ex. 1

Explain why the following code doesn’t create a circular list.

```{r}
x <- list()
x[[1]] <- x
```

### Answer

Because the 1st `x` refers to a different object than the 2nd `x` altogether. It's not a loop. Upon the second command, a new list is created, with its first element pointing to the list behind the first `x`, then the new list is bound to `x`.

```{r}
# Profile the 1st x
x <- list()
lobstr::ref(x)
lobstr::tree(x)

# Profile the 2nd x
x[[1]] <- x
lobstr::ref(x)
lobstr::tree(x)
```

## Ex. 2

Wrap the two methods for subtracting medians into two functions, then use the `bench` package to carefully compare their speeds. How does performance change as the number of columns increase?

**NB:** The methods are:

```{r}
# Subtract median from all elements of a dataframe
method_1 <- function(x, medians) {
  for (i in seq_along(medians)) {
    x[[i]] <- x[[i]] - medians[[i]]
  }
}

# Coerce dataframe to list prior to calculation of median
method_2 <- function(x, medians) {
  y <- as.list(x)

  for (i in seq_along(medians)) {
    y[[i]] <- y[[i]] - medians[[i]]
  }
}
```

### Answer

Method 2 would be faster than method 1 since fewer copies are performed.

```{r}
# For the following dataframe and medians
x <- data.frame(matrix(
  runif(5 * 1e4),
  ncol = 5
))

medians <- vapply(
  x,
  median,
  numeric(1)
)

# Compare performance
bench::mark(
  method_1(x, medians),
  method_2(x, medians)
)
```

The 2nd method seems to be 3x faster (likely due to needing fewer copies)

```{r}
# For the following dataframe and medians
x <- data.frame(matrix(
  runif(500 * 1e4),
  ncol = 500
))

medians <- vapply(
  x,
  median,
  numeric(1)
)

# Compare performance
bench::mark(
  method_1(x, medians),
  method_2(x, medians)
)
```

## Ex. 3

What happens if you attempt to use tracemem() on an environment?

### Answer

Experiment:

```{r}
#| error: TRUE

env_test <- rlang::env(
  a = 1,
  b = 2,
  c = 3
)

tracemem(env_test)
```

`tracemem()` will not trace it, because environments are not copied when they are modified. They are always modified in place.