---
format: live-html
webr:
  packages:
    - here
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

# Control flow

```{r setup}
#| include: FALSE

# Declare location
here::i_am("quarto/4_chapter5.qmd")

# Import packages
library("lobstr")
library("tidyverse")

# Set seed
set.seed(67)
```

# Quiz

## Quiz 1

What is the difference between `if` and `ifelse()`?

### Answers

The difference between these conditions are:
- `if()` operates on a single logical value
- `ifelse()` operates on a vector of logical values. Syntax is also different.

```{r}
#| eval: false
?Control
?ifelse
```

```{r}
# if
x <- TRUE
if (x) {
  print("Hello")
}

# ifelse
x <- c(TRUE, FALSE, TRUE, TRUE)
ifelse(
  x,
  print("Hello"),
  print("Goodbye")
)
```

## Quiz 2

In the following code, what will the value of `y` be if `x` is `TRUE`? What if `x` is `FALSE`? What if `x` is `NA`?

```{r}
#| eval: false
y <- if (x) 3
```

### Answers

Outcome of all cases are:
- `x <- TRUE`: `y` is assigned.
- `x <- FALSE`: `y` is NULL
- `x <- NA`: Error.

```{r}
#| error: true
# Let's test
## x is TRUE
x <- TRUE
y <- if (x) 3
y
## x is FALSE
x <- FALSE
y <- if (x) 3
y
## x is NA
x <- NA
y <- if (x) 3
y
```

## Quiz 3

What does `switch("x", x = , y = 2, z = 3)` return?

### Answers

`switch` evaluates the matched element or the first following non-missing argument. Since `x` is missing, the next argument `y` is evaluated, yielding `2`.

```{r}
switch("x", x = , y = 2, z = 3)
```

# Choices

## Ex. 1

What type of vector does each of the following calls to `ifelse()` return?

```{r}
#| eval: false
ifelse(TRUE, 1, "no")
ifelse(FALSE, 1, "no")
ifelse(NA, 1, "no")
```

### Answers

```{r}
# Evaluates to the TRUE arg, 1
ifelse(TRUE, 1, "no")

# Evaluates to the FALSE arg, no
ifelse(FALSE, 1, "no")

# NA propagates, result is NA
ifelse(NA, 1, "no")
```


## Ex. 2

Why does the following code work?

```{r}
x <- 1:10
if (length(x)) {
  "not empty"
} else {
  "empty"
}

x <- numeric()
if (length(x)) {
  "not empty"
} else {
  "empty"
}
```

### Answers

The code works because:

- For the vector `1:10`, its length is 10 and is equal to `TRUE`.
- For the vector `numeric()`, its length is 10 and is equal to `FALSE`.

```{r}
# Length of x is 10, equal to TRUE
x <- 1:10
length(x)
length(x) == TRUE

# Length of x is 0, equal to FALSE
x <- numeric()
length(x)
length(x) == FALSE
```

# Loops

## Ex. 1

Why does this code succeed without errors or warnings?

```{r}
x <- numeric()
out <- vector("list", length(x))
for (i in 1:length(x)) {
  out[i] <- x[i]^2
}
out
```

### Answers

This horrorshow works because:

- `vector()` by default, creates an empty vector. This behaviour is intended.
- `1:length(x)` is a vector counting down from 1 to 0 (`c(1, 0)`)

```{r}
# Example case
x1 <- vector(
  mode = "list",
  length = numeric() %>%
    length()
)

# Default case
x2 <- vector(
  mode = "list"
)

identical(x1, x2)

```

During the first iteration, the following occurs:

- The subsetting of `x` successfully returns `NA` as intended.
- The `out[1]` list element is successfully set as intended.

```{r}
x <- numeric()
out <- vector("list", length(x))

out[1] <- x[1]^2
```

During the second iteration, the following occurs:

- The subsetting of `x` returns a zero length vector not containing any data.
- The subsetting of `out` returns a zero length vector not containing any data.
- No data accessed or modified.

```{r}
x <- numeric()
out <- vector("list", length(x))

out[0] <- x[0]^2
```

In summary, all operations behaved as intended by the code and contained no errors. 

## Ex. 2

When the following code is evaluated, what can you say about the vector being iterated?

```{r}
xs <- c(1, 2, 3)
for (x in xs) {
  xs <- c(xs, x * 2)
}
xs
```

### Answers

In every iteration, the calculated value is concatenated with the previous vector and attached to the end. The `for` looping section is evaluated prior to beginning the loop, which ensures that the loop ends in 3 iterations, even though the length of `xs` is increased by the loop. 

```{r}
xs <- c(1, 2, 3)
for (x in xs) {
  xs <- c(xs, x * 2)
  print(xs)
}
```

## Ex. 3

```{r}
for (i in 1:3) {
  i <- i * 2
  print(i)
}
```

### Answers

In essence, there are two different `i` variables here:

- The first `i` is outside of the loop and is responsible for counting the number of iterations (`1, 2, 3`)
- The second `i` is inside the loop.

At every iteration, the outer `i` is passed in the loop, doubled, printed, then replaced with the outer `i` when the next iteration begins. The outer `i` is never modified.